Relationships in the extremes and their influence on competition and coexistence

### 2022-03-11 paper/Paper.Rnw
a few notes about progress editing were added
edited first para of Discussion

%G4
To achieve goal G4 from the Introduction, i.e., to provide an empirical example of species 
coexistence which highlights the role ATAs can play, we use a laboratory chemostat
system of freshwater diatoms explored by \cite{Descamps_2005}.
Those authors used measurements of growth rates of \emph{Fragilaria crotonensis} and 
\emph{Cyclotella pseudostelligera} over a range of temperatures to parametrize a resource
competition model. The model is 
\begin{align}
\frac{dS}{dt} &= D(S_0 - S) - Q_1x_1 \frac{V_1 S}{K_1 + S} - Q_2 x_2 \frac{V_2 S}{K_2 + S} \\
\frac{dx_j}{dt} &= x_j \frac{V_j S}{K_j+S}-D x_j, 
\end{align}
for $j=1,2$, where $S$ is a extracellular silicate concentration in the chemostat, $x_1$ is the population density
of \emph{F. crotonensis} and $x_2$ is that of \emph{C. pseudostelligera}, 
$D$ is the chemostat outflow rate, $S_0$ is concentration of silicates in the chemostat
inflow, the $Q_j$ are amounts of silicate per cell, the $V_j$ are maximum population growth rates, and
the $K_j$ are half-saturation constants for the dependence of growth rates on nutrient concentration.
The temperature-dependent parameters $Q_j$, $V_j$, and $K_j$ were obtained by 
\cite{Descamps_2005} through batch experiments and curve fitting or interpolation.
\cite{Descamps_2005} showed that permanent coexistence of the two species occurs when 
temperature fluctuates periodically, 
$\theta(t)=\theta_0 +  a\sin(2\pi t/P)$, 
with parameters $\theta_0 = 18^\circ$C, $a = 6^\circ$C, $P = 60$d. 
\cite{Ellner_2016, Ellner_2019} broke down coexistence into contributing mechanisms
according to (\ref{eq:previous_decomp}) for these same values of
$\theta_0$, $a$ and $P$. We further decomponse
coexistence according to (\ref{eq:full_decomp}), and we do so for ranges of values
of $\theta_0$, $a$ and $P$. In contrast to the lottery model case, for which simplifying assumptions
made it sufficient to decompose only the GWR of species 1, for this
%***DAN: Please search for "growth when rare" and replace with GWR (except for the first instance,
%where GWR is defined). If one defines an acronym, one then has to use it.
example we considered $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$ for both $i=1$, $j=2$ and for
$i=2$, $j=1$. Further details of the model setup are in SI section \ref{SI-sect:diatom_model}
and specifics of how our decomposition extends to this empirical case are in 
SI section \ref{SI-sect:Empircal_app}.
%***DAN: Jasmin, you had already created and started on the second of these SI sections, but the first one
%is new, and needs to be filled in. Please make a start at that when you get a chance. I've put some 
%notes there indicating what the content of that section should be.
%***JASMIN: Working on filling out diatom model set up section

%***DAN: I see many instances of IGR have been replaced by GWR, but there are some IGRs remaining, e.g., in the sup mat, and
%on the figures.

@@

We produced and applied a mathematical extension to MCT that, for the first time, 
considers ATAs for the coexistence of species and maintenance of biodiversity. 
Implementing new methods of MCT \citep{Ellner_2016, Ellner_2019}, we provided a further tertiary 
decomposition to MCT and decompose the storage effect into two new terms, 
$\Delta_i^{[EC]}$, the contribution of ATAs, and $\Delta_i^{[E \nat C]}$, the contribution of correlation \emph{per se}. 
The results of our application to a simple lottery model and 
to a competitive diatom model show that ATA contributions to storage effects and to coexistence
are non-negligible and meaningfully contribute to the coexistence of species and thus the maintenance of biodiversity. 
Our results reinforce that common methods of correlation fail to reflect and include 
important information on the structure of the relationship between variables and amends
coexistence to the list of ecological theory and phenomena that ATAs have been found to 
be important for \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of covariance between drivers to coexistence, classically \emph{EC covariance},
\citep{Chesson_2012, Ellner_2016, Ellner_2019}, we suggest that storage effects, instead, 
be described as the contribution of the \emph{relationship} between the drivers.
Measures of covariance do not fully capture all aspects of associativity throughout the bivariate 
distribution of the variables; thus, it may be misleading to describe the contribution of the whole 
EC (or other drivers') relationship as the contribution of EC (or other drivers') covariance.
Accordingly, we have separated the mechanisms of storage effect 
into contributions of ATAs and contributions of correlation \emph{per se}.

>>>>

%G4
To achieve goal G4 from the Introduction, i.e., to provide an empirical example of species 
coexistence which highlights the role ATAs can play, we use a laboratory chemostat
system of freshwater diatoms explored by \cite{Descamps_2005}.
Those authors used measurements of growth rates of \emph{Fragilaria crotonensis} and 
\emph{Cyclotella pseudostelligera} over a range of temperatures to parametrize a resource
competition model. The model is 
\begin{align}
\frac{dS}{dt} &= D(S_0 - S) - Q_1x_1 \frac{V_1 S}{K_1 + S} - Q_2 x_2 \frac{V_2 S}{K_2 + S} \\
\frac{dx_j}{dt} &= x_j \frac{V_j S}{K_j+S}-D x_j, 
\end{align}
for $j=1,2$, where $S$ is a extracellular silicate concentration in the chemostat, $x_1$ is the population density
of \emph{F. crotonensis} and $x_2$ is that of \emph{C. pseudostelligera}, 
$D$ is the chemostat outflow rate, $S_0$ is concentration of silicates in the chemostat
inflow, the $Q_j$ are amounts of silicate per cell, the $V_j$ are maximum population growth rates, and
the $K_j$ are half-saturation constants for the dependence of growth rates on nutrient concentration.
The temperature-dependent parameters $Q_j$, $V_j$, and $K_j$ were obtained by 
\cite{Descamps_2005} through batch experiments and curve fitting or interpolation.
\cite{Descamps_2005} showed that permanent coexistence of the two species occurs when 
temperature fluctuates periodically, 
$\theta(t)=\theta_0 +  a\sin(2\pi t/P)$, 
with parameters $\theta_0 = 18^\circ$C, $a = 6^\circ$C, $P = 60$d. 
\cite{Ellner_2016, Ellner_2019} broke down coexistence into contributing mechanisms
according to (\ref{eq:previous_decomp}) for these same values of
$\theta_0$, $a$ and $P$. We further decomponse
coexistence according to (\ref{eq:full_decomp}), and we do so for ranges of values
of $\theta_0$, $a$ and $P$. In contrast to the lottery model case, for which simplifying assumptions
made it sufficient to decompose only the GWR of species 1, for this
%***DAN: Please search for "growth when rare" and replace with GWR (except for the first instance,
%where GWR is defined). If one defines an acronym, one then has to use it.
example we considered $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$ for both $i=1$, $j=2$ and for
$i=2$, $j=1$. Further details of the model setup are in SI section \ref{SI-sect:diatom_model}
and specifics of how our decomposition extends to this empirical case are in 
SI section \ref{SI-sect:Empircal_app}.

%***DAN: I see many instances of IGR have been replaced by GWR, but there are some IGRs remaining, e.g., 
%on the figures.

@@

%***DAN: Jasmin, I just want to point out the below paragraph was quite well structured as it was 
%before I edited it. I only made changes on the sentence level. Nice job! Go aead and delete this 
%comment once you read it.
We produced and applied a new extension to MCT that 
quantifies the impact of ATAs on species coexistence and the maintenance of biodiversity. 
Building upon a recent, simulation-based approach to MCT \citep{Ellner_2016, Ellner_2019},
we decomposed storage effects into two new mechanisms, the contribution of ATAs to coexistence,
and the contribution of $EC$ correlation \emph{per se}.
Applications of our extended theory to the lottery model and to models based on a laboratory 
system of competing diatoms demonstrated that  
ATA contributions to storage effects and to species coexistence and the maintenance of biodiversity
can often be comparable in magnitude to the contributions of previously recognized mechanisms.
Our results highlight that common methods of correlation and covariance fail to include 
important information on the structure of relationships between variables, and our results add
coexistence and biodiversity maintenance to the list of ecological theories and phenomena on which 
ATAs have been shown to have an important impact \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of $EC$ \emph{covariance} to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019},
our results suggest that storage effects should be redefined to constitute the 
contribution of \emph{associations} between $E$ and $C$, including both ATA effects and
correlation \emph{per se}; the prior terminology, \emph{covariance}, is typically construed 
as reflecting the linear relationship between variables, ignoring complexities of the association
such as ATAs.

%***DAN: Dan made it to here in editing and has to stop until after spring break
